---
title: "Exercise 11"
output: html_document
---
Keerthana Stanley


ANOVA Practice


loading in data:
```{r}
library(tidyverse)

f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/AVONETdataset1.csv"

d <- read_csv(f, col_names = TRUE)

```



Winnow the dataset to include only the following variables: 
Species1, Family1, Order1, Beak.Length_Culmen, Beak.Width, Beak.Depth, Tarsus.Length, Wing.Length, Tail.Length, Mass, Habitat, Migration, Trophic.Level, Trophic.Niche, Min.Latitude, Max.Latitude, Centroid.Latitude, Primary.Lifestyle, ** and Range.Size


use select() to choose specific variables
also using the pipe operator here
```{r}
d_winnowed <- d %>% 
  select(Species1, Family1, Order1, Beak.Length_Culmen, Beak.Width, Beak.Depth, Tarsus.Length, Wing.Length, Tail.Length, Mass, Habitat, Migration, Trophic.Level, Trophic.Niche, Min.Latitude, Max.Latitude, Centroid.Latitude, Primary.Lifestyle, Range.Size)
```




Do a bit of exploratory data analysis with this dataset, e.g., using the {skimr} package. 


QUESTION:
Which of the variables are categorical and which are numeric?


ANSWER:
by looking under the skim_type, I can see which variables are categorical (character) versus numeric

  categorical: Species1, Family1, Order1, Habitat, Trophic.Level, Trophic.Niche, Primary.Lifestyle

  numeric: Beak.Length_Culmen, Beak.Width, Beak.Depth, Tarsus.Length, Wing.Length, Tail.Length, Mass,                     Migration, Min.Latitude, Max.Latitude, Centroid.Latitude, Range.Size
```{r}
library(skimr)

d_sumstats <- skim(d_winnowed)
```


_____________________________________________________________________________________________________________
CHALLENGE 1: One-Factor ANOVA

STEP 1-->

BOXPLOTS--> of log(Mass) in relation to Trophic.Level and Migration behavior type

For each plot, drop from the visualization all species records where the categorical variable of interest is missing from the dataset. 

*** convert the variable Migration (which is scored as a number: “1”, “2”, or “3”) from class numeric to either being classified as a factor or as a character (string) variable!



CONVERTING MIGRATION TO BE CATEGORICAL
can use the factor() function to do so:
  - categorical variables are represented as FACTORS, not strings!!
  - reference to Module 20
more documentation- https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/factor
```{r}
d_winnowed$Migration <- factor(d_winnowed$Migration, levels = c("1", "2", "3"))

# re-running this to check
d_sumstats <- skim(d_winnowed)
```
now we can see that Migration is classified as a factor, not numeric

FILTERING DATA TO DROP MISSING DATA!
using filter()
```{r}
d_filtered <- d_winnowed %>%
  filter(!is.na(Trophic.Level) & !is.na(Migration)) #drops missing values for Migration and Trophic Level
```
using !is.na to remove missing values:
https://stackoverflow.com/questions/28857653/removing-na-observations-with-dplyrfilter



we already have Mass but we need log(Mass)

I'm using mutate() to add a new column for this!
```{r}
d_filtered <- d_filtered %>%
  mutate(log_Mass = log(Mass))
```

now time to make our boxplots-- using ggplot2

also using geom_jitter() as shown in class- this function allows me to adjust (a.k.a 'jitter') points to prevent overplotting and clearly see the distributions of the data (need to adjust alpha value to make these points low opacity)

documentation- https://ggplot2.tidyverse.org/reference/geom_jitter.html#:~:text=The%20jitter%20geom%20is%20a,by%20discreteness%20in%20smaller%20datasets.
```{r}
library(ggplot2)

# where x=log(Mass) and y=Tropic Level
ggplot(data = d_filtered, aes(y = log_Mass, x = Trophic.Level)) + geom_boxplot() + geom_jitter(position = "jitter", alpha = 0.03)

# where x=log(Mass) and y= Migration
ggplot(data = d_filtered, aes(y = log_Mass, x = Migration)) + geom_boxplot() + geom_jitter(position = "jitter", alpha = 0.03)
```

----------------------------------
STEP 2-->

Run linear models using the lm() function to look at the relationship between log(Mass) and Trophic.Level and between log(Mass) and Migration.

QUESTIONS:

  - Examine the output of the resultant linear models. Is log(Mass) associated with either Trophic.Level or Migration category? That is, in the global test of significance, is the F statistic large enough to reject the null hypothesis of an F value of zero?

  - Given the regression coefficients returned for your Migration model, which Migration categor(ies) are different than the reference level? 
  - What level is the reference level? 
          - Re-level and assess differences among the remaining pair of Migration categories.


ANSWERS:

  - As we learned in Module 19, we can determine the significance based on the p-value of the F-statistic...the p value of the F statistic for BOTH Trophic Level and Migration is 2.2e-16. at such as small level we MUST REJECT THE NULL HYPOTHESIS. Additionally when calculating the critical value for m1 and m2 using the qf() function, both F-stat values are well above their respective critical values
  
  - The regression coefficient levels for each migration are as follows. As we can see, Migration 1 is our reference, and Migrations 2 and 3 are both quite different from it
(Intercept)             3.77457 
Migration2              0.75971  
Migration3              0.37647 

  - POST-RELEVELING: here we see that when 2 or 3 become the reference, it's much easier to distinguish the differences between them (the releveling code is shown below!)
-------
REF = 2
(Intercept)             4.53428 
Migration3              -0.38324
Migration1              -0.75971
-------
REF = 3
(Intercept)             4.15104  
Migration2              0.38324 
Migration1              -0.37647
---------------------------------


As shown in Modules 18 and 20, the lm() function is pretty simple to use here:

m1--> relation between log(Mass) and Tropic Level
```{r}
m1 <- lm(data = d_filtered, log_Mass ~ Trophic.Level)

summary(m1)
```

```{r}
(critical_val_m1 <- qf(p = 0.95, df1 = 3, df2 = 10982))
# where alpha = 0.05 and p = 1 - 0.05
```





m2--> relation between log(Mass) and Migration
```{r}
m2 <- lm(data = d_filtered, log_Mass ~ Migration)

summary(m2)
```

```{r}
(critical_val_m2 <- qf(p = 0.95, df1 = 2, df2 = 10983))
```


now let's use relevel() function for Migration, as shown in Module 20
```{r}
d_filtered$Migration <- relevel(d_filtered$Migration, ref = "2") # assuming migratory type 2 to be baseline now
m2_relevel <- lm(log_Mass ~ Migration, data = d_filtered)
summary(m2_relevel)
```
-----------------------------------------------
STEP 3-->

Conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which Migration categories differ “significantly” from one another 


Module 20 states that the TukeyHSD() function is run on the output of the ANOVA (aov()) function

thus I need to first use aov()
```{r}
(m2_anova <- aov(log_Mass ~ Migration, data = d_filtered))
```

```{r}
(posthoc <- TukeyHSD(m2_anova, which = "Migration", ordered = TRUE, conf.level = 0.95))
```
now I should pass the Tukey test output to a plot() function
```{r}
plot(posthoc, xlim = c(0, 1)) # started with wider x-limits then narrowed it down to make sure it contains all the data
```
if the data was not significantly different between any of the migration types, the confidence intervals would cross zero (indicating that there is potentially zero difference in the means). however, we see that this is not the case, therefore we can confirm that each migration type IS SIGNIFICANTLY different from one another

---------------------------------------------------
STEP 4
- Use a permutation approach
- Inference to generate a null distribution of F statistic values for the model of log(Mass) in relation to Trophic.Level and calculate a p value for your original F statistic. 

Do this either by:
  1. programming your own permutation test (e.g., by shuffling values for the predictor or response variable among observations and calculating an F statistic for each replicate) 
  
  or 
  
  2. by using the {infer} workflow and setting calculate(stat="F").


------
before I can compute permuted F-statistic values, I must compute the ACTUAL/original F-stat for log_Mass and Tropic.Level

r documentation shows that the calculate() function from {infer} workflow can compute the F-stat easily-
https://www.rdocumentation.org/packages/infer/versions/1.0.5/topics/calculate

(similar use of the functions I will be using for permuted f stats)

```{r}
library(infer)

(original.F <- d_filtered %>%
  specify(log_Mass ~ Trophic.Level) %>%
  calculate(stat = "F"))
```


now I can follow our methology from Module 16 (or more specific to this, Module 20.5) and generate a permutation test
```{r}
permuted.F <- d_filtered %>%
  specify(log_Mass ~ Trophic.Level) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "F")


# plot
visualize(permuted.F) +
  shade_p_value(obs_stat=original.F, direction="greater")
```
get p-value
```{r}
(p.value <- permuted.F %>%
  get_p_value(obs_stat = original.F, direction="greater"))
```


_______________________________________________________________________________________________________
CHALLENGE 2:



STEP 1-->

Create the following two new variables and add them to AVONET dataset:

  Relative beak length, which you should calculate as the residual of log(Beak.Length_Culmen) on log(Mass).

  Relative tarsus length, which you should calculate as the residual of log(Tarsus.Length) on log(Mass).
  
first I'm geting the logs of the variables I need

```{r}
d_filtered <- d_filtered %>%  
  mutate(log_Beak = log(Beak.Length_Culmen),
         log_Tarsus = log(Tarsus.Length)) %>%
  mutate(rel_Beak = residuals(lm(log_Beak ~ log_Mass, data = .)),
         rel_Tarsus = residuals(lm(log_Tarsus ~ log_Mass, data = .))) %>%
    filter(!is.na(rel_Tarsus) & 
         !is.na(Primary.Lifestyle) & 
         !is.na(rel_Beak) & 
         !is.na(Trophic.Niche))

```


------------------------------------------------------
STEP 2-->

Make a boxplot or violin plot of your new relative tarsus length variable in relation to Primary.Lifestyle and of your new relative beak length variable in relation to Trophic.Niche


I'm choosing to make boxplots similar to Challenge 1
catgorical --> x; numeric --> y
rel_Tarsus (y) and Primary.Lifestyle (x)
rel_Beak (y) and Trophic.Niche (x)

```{r}
# rel_Tarsus (y) and Primary.Lifestyle (x)
ggplot(data = d_filtered, aes(y = rel_Tarsus, x = Primary.Lifestyle)) + geom_boxplot() + geom_jitter(position = "jitter", alpha = 0.03)

# rel_Beak (y) and Trophic.Niche (x)
# adjust angle of x-ticks to better see each word
ggplot(data = d_filtered, aes(y = rel_Beak, x = Trophic.Niche)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 45, hjust = 1))+ geom_jitter(position = "jitter", alpha = 0.03) 
```

-----------------------------------------------------
STEP 3 -->

Run ANOVA analyses to look at the association between geographic range size and the variable Migration. 

- drop those observations for which Migration is not scored
- look at the distribution of the variable Range.Size
    - decide whether and how it might need to be transformed. 

QUESTIONS:    
  - Based on the global model, is range size associated with form of migration? 
  - How much of the variance in your measure of range size is associated with Migration behavior style?
  - Given the regression coefficients returned in the output of the model, which Migration categor(ies) are different than the reference level? 
  - What level is the reference level? Relevel and assess differences among the remaining pair of Migration categories. 

Also conduct a post-hoc Tukey Honest Significant Differences test to also evaluate which Migration categories differ “significantly” from one another (see Module 20).


ANSWERS:
  - the ANOVA results show a very high F-statistic value (above the critical value) with a very low p-value, suggesting a significant association between range size and migration
  - the R-squared value from the linear model summary tells us what percentage of the variance in Range.Size is associated with Migration behavior; from my linear model I got 0.08681 or around 8.7% (which honestly surprised me a little, I thought it would be higher?)
  - The reference level is Migration type 2, and both types 1 and 3 seem to be different from it
  - AFTER RELEVELING: 1 and 3 also seem to be more different from one another
  - none of the CI's from the Tukey test cross over zero, meaning that all three types of migration differ significantly from one another

already dropped missing values for Migration
x = Migration, y = Range.Size

checking distribution of Range.Size first using ggplot2
```{r}
ggplot(d_filtered, aes(x = Range.Size)) +
  geom_histogram(bins = 10)
```
the values for Range Size are superrrr small, so that indicates to my that I should use ther log of it
```{r}
log_RangeSize <- log(d_filtered$Range.Size)

d_filtered <- d_filtered %>%
  mutate(log_RangeSize)
```

now I'm ready for running ANOVA analysis using the aov() function
```{r}
m3_anova <- aov(log_RangeSize ~ Migration, data = d_filtered)

summary(m3_anova)
```
could also find the f-stat the same way as I did in Challenge 1
```{r}
m3 <-lm(log_RangeSize ~ Migration, data = d_filtered)
summary(m3)
```
```{r}
(crit <- qf(p = 0.95, df1 = 1, df2 = 10929))
```
the F-stat is wellll above this critical value

Releveling:
```{r}
d_filtered$Migration <- relevel(d_filtered$Migration, ref = "1") # assuming migratory type 2 to be baseline now
m3_relevel <- lm(log_RangeSize ~ Migration, data = d_filtered)
summary(m3_relevel)
```
```{r}
d_filtered$Migration <- relevel(d_filtered$Migration, ref = "3") # assuming migratory type 2 to be baseline now
m3_relevel <- lm(log_RangeSize ~ Migration, data = d_filtered)
summary(m3_relevel)
```

conducting a post-hoc Tukey Honest Significant Differences test
```{r}
(posthoc <- TukeyHSD(m3_anova, which = "Migration", ordered = TRUE, conf.level = 0.95))
```
```{r}
plot(posthoc, xlim = c(0,4))
```
none of the CI's cross 0!! so there is a signifcant difference between each of the migration types

--------------------------------------------
STEP 4-->

Winnow your original data to just consider birds from the Infraorder “Passeriformes” (song birds).

Run separate one-factor ANOVA analyses to look at the association between 
[1] relative beak length and Primary.Lifestyle 
[2] relative beak length and Trophic.Level. 

Make boxplots of response variable by each predictor and by the combination of predictors.

Run linear models for each predictor separately and interpret the model output.

winnowing the data to ONLY consider passeriformes
```{r}
passeriformes <- d_filtered %>%
  filter(Order1 == "Passeriformes")
```


run anova b/t rel_Beak and Primary.Lifestyle, then rel_Beak and Trophic.Level
```{r}
# m4- rel_Beak (y) and Primary.Lifestyle (x)
m4_anova <- aov(rel_Beak ~ Primary.Lifestyle, data = passeriformes)
summary(m4_anova)
```

```{r}
# m5- rel_Beak (y) and Trophic.Level (x)
m5_anova <- aov(rel_Beak ~ Trophic.Level, data = passeriformes)
summary(m5_anova)
```

make boxplots of response variable by predictor AND combination of predictors

for individual predictors
```{r}
# rel_Beak (y) and Primary.Lifestyle (x)
ggplot(data = passeriformes, aes(y = rel_Beak, x = Primary.Lifestyle)) + geom_boxplot() + geom_jitter(position = "jitter", alpha = 0.03)

# rel_Beak (y) and Trophic.Level (x)
# adjust angle of x-ticks to better see each word
ggplot(data = passeriformes, aes(y = rel_Beak, x = Trophic.Level)) + geom_boxplot() + geom_jitter(position = "jitter", alpha = 0.03) 
```


combined predictors: I looked at Module 20.8

uses facet_wrap() function
```{r}
# rel_Beak (y) and Primary.Lifestyle (x)
# Trophic.Level is facet wrapped
ggplot(data = passeriformes, aes(y = rel_Beak, x = Primary.Lifestyle)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + #needed to rotate x-labels to fit in the full text
  facet_wrap(~ Trophic.Level, ncol=3) + # 3 levels (carnivore, herbivore, omnivore)
  xlab("Primary Lifestyle") + ylab("Relative Beak Length") + geom_jitter(position = "jitter", alpha = 0.03)

# rel_Beak (y) and Trophic.Level (x)
# Primary.Lifestyle is facet wrapped
ggplot(data = passeriformes, aes(y = rel_Beak, x = Trophic.Level)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + #needed to rotate x-labels to fit in the full text
  facet_wrap(~ Primary.Lifestyle, ncol=4) + # 4 levels (aerial, generalist, insessorial, terrestrial)
  xlab("Trophic Level") + ylab("Relative Beak Length") + geom_jitter(position = "jitter", alpha = 0.03)
```
run linear models (lm) b/t rel_Beak and Primary.Lifestyle, then rel_Beak and Trophic.Level
```{r}
# m4- rel_Beak (y) and Primary.Lifestyle (x)
m4_lm <- lm(rel_Beak ~ Primary.Lifestyle, data = passeriformes)
summary(m4_lm)
```

```{r}
# m5- rel_Beak (y) and Trophic.Level (x)
m5_lm <- lm(rel_Beak ~ Trophic.Level, data = passeriformes)
summary(m5_lm)
```
from the results of the linear modeling/ANOVA, we can see that both of the F-stats are high enough (and their p-values low enough) to suggest a significant relationship between relative beak length and BOTH trophic level and primary lifestyle in passeriformes. based on the R^2 values we also see there percentage of variance that is significantly related is around 5% for both

--------------------------------------------------------
STEP 5-->

Run a two-factor model to look at the association between relative beak length and both Primary.Lifestyle and Trophic.Level among the passeriformes. 

QUESTION:
  - Based on the model output, what would you conclude about how relative beak length is related to these two variables?
  
ANSWER:
  - once again, I see that the F-stat/p-value results from the linear model summary does indicate significant relation between beak length and both primary lifestyle AND trophic level. most notably I see that R^2 has increased (actually doubled) to around 11% of beak length variance that is related to and explained by the model. these suggests suggest that beak length is influenced by multiple factors, including both primary lifestyle and trophic level


I looked at Module 20.8 with Multiple Factor ANOVA
```{r}
(m_combined <- summary(lm(data = passeriformes, rel_Beak ~ Primary.Lifestyle + Trophic.Level)))
```
-----------------------------------------------------
STEP 6-->

Finally, run an additional two-way model with the same dataset and predictors, but adding the possibility of an interaction term. 

To do this, you should modify your model formula using the colon operator (:) to specify the interaction, e.g., relative beak length ~ Primary.Lifestyle + Trophic.Level + Primary.Lifestyle:Trophic.Level. 

QUESTION:
  - Based on the model output, what would you now conclude about how relative beak length is related to these two variables?
  
ANSWER:
  - looking at the p-value for the F-stat in "Primary.lifestyle:Trophic.Level" interactions row, the interaction between these two variables IS SIGNIFICANT. this suggests that primary lifestyle and trophic level are somewhat codepedent upon one another, in addition to influencing beak length. 
  
once again I'm referencing to Module 20.8 to use the colon operator
```{r}
m <- aov(data = passeriformes, rel_Beak ~ Primary.Lifestyle + Trophic.Level + Primary.Lifestyle:Trophic.Level)
# the colon (:) operator includes specific interaction terms
summary(m)
```

--------------------------------------------------------
STEP 7-->

Use the interaction.plot() function to visualize the interaction between Primary.Lifestyle and Trophic.Level 

once again, the code for this is influenced by the instructions in Module 20.8


the x factor is primary lifestyle and trace factor is trophic level
```{r}
interaction.plot(
  x.factor = passeriformes$Primary.Lifestyle,
  xlab = "primary lifestyle",
  trace.factor = passeriformes$Trophic.Level,
  trace.label = "trophic level",
  response = passeriformes$rel_Beak,
  fun = base::mean, # make sure we use {base} version
  ylab = "relative beak length"
)
```

just switching the x factor to trophic level and trace to primary lifestyle
```{r}
interaction.plot(
  trace.factor = passeriformes$Primary.Lifestyle,
  trace.label = "primary lifestyle",
  x.factor = passeriformes$Trophic.Level,
  xlab = "trophic level",
  response = passeriformes$rel_Beak,
  fun = base::mean, # make sure we use {base} version
  ylab = "relative beak length"
)
```
























